---
title: "PA1_template.Rmd"
author: "William P Fawcett"
date: "February 28, 2016"
output: html_document
---

```{r echo = TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(lattice)
```
Download data files and unzip.
```{r echo = TRUE}
myURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(myURL, destfile = "activity.zip")
Download_date <- Sys.time()
unzip("activity.zip")
```
Read in the data file: "activity.csv"
```{r echo = TRUE}
Activity_dataset <- read.csv("activity.csv")
Activity_dataset <- tbl_df(Activity_dataset)
dim(Activity_dataset)
Activity_dataset
```
Determine the total number of steps taken per day.
Plot histogram.
Calculate the mean and median number of the total number of steps taken per day.
```{r echo = TRUE}
Total_Steps <- Activity_dataset %>% group_by(date) %>% 
  summarize(Total = sum(steps, na.rm = TRUE))
Total_Steps
hist(Total_Steps$Total, main = "Total Steps Per Day", xlab = "Total Steps", 
     col = "red", xlim = c(0,22000))
summary(Total_Steps$Total)
Mean_Total_Steps <- formatC(mean(Total_Steps$Total, na.rm = TRUE), digits = 3, format = "f")
Median_Total_Steps <- formatC(median(Total_Steps$Total, na.rm = TRUE), digits =3, format = "f")
Mean_Total_Steps
Median_Total_Steps
```
Explore the average daily activity pattern.
Calculate the mean for each time interval (288 intervals).
Determine the interval that has the maximum number of steps.
```{r echo = TRUE}
Activity_dataset <- ungroup(Activity_dataset)
Interval <- Activity_dataset %>% group_by(interval) %>% 
summarize(avg_steps = formatC(mean(steps, na.rm = TRUE), digits = 3, format = "f"))
Interval$avg_steps <- as.numeric(Interval$avg_steps)
Interval
```
Generate an XY line plot.
```{r echo = TRUE}
plot(Interval$interval,Interval$avg_steps, type="l",
     xlab = "Inverval", ylab = "Number of Steps", 
     main = "Average Number of Steps by Interval")
```
Determine the interval that has the maximum number of steps.
```{r echo = TRUE}
MaxStepInterval <- as.numeric(filter(Interval, avg_steps == max(avg_steps)) %>%
  select(interval))
MaxStepInterval
```
Calculate the number of rows with "NA"
```{r echo = TRUE}
Missing_Step_Values <- nrow(filter(Activity_dataset, is.na(steps)))
Missing_Step_Values
```
Create a new dataset similar to Activity_dataset.
Impute missing step values: Use mean value for each interval.
```{r echo = TRUE}
Filled_In <- Activity_dataset
for(i in 1:17568){
  if (is.na(Filled_In$steps[i])){
    x <- Filled_In$interval[i]
    j <- 1
    for (j in 1:288){
      if(x == Interval$interval[j]){
        Filled_In$steps[i] <- Interval$avg_steps[j]
      }
    }
  }
}
Filled_In$steps <- as.numeric(Filled_In$steps)
Filled_In
```
Generate histogram for New Total Steps.
```{r echo = TRUE}
New_Total_Steps <- Filled_In %>% 
  group_by(date) %>% 
  summarize(Total = sum(steps, na.rm = TRUE))
hist(New_Total_Steps$Total, main = "Total Steps Per Day", xlab = "Total Steps", 
     col = "red", xlim = c(0,22000))
New_Mean_Total_Steps <-formatC(mean(New_Total_Steps$Total, na.rm = TRUE), digits = 3, format = "f") 
New_Mean_Total_Steps
New_Median_Total_Steps <- formatC(median(New_Total_Steps$Total, na.rm = TRUE), digits = 3, format = "f")
New_Median_Total_Steps
```
Compare New Mean/Median Values with the old (original) Mean/Median Values.
Use a Boxplot. As can be seen the New Mean/Median values (Filled_In) are higher than the original values. The dash lines are the mean values for Missing_Values (red) and Filled_In (blue).
```{r echo = TRUE}
CompareTotals <- cbind(Total_Steps$Total, New_Total_Steps$Total)
colnames(CompareTotals) <- c("Missing_Values", "Filled_In")
boxplot(CompareTotals, col = c("red", "blue"), ylab = "Total Steps")
abline(h=Mean_Total_Steps, col = "red", lty = 2, lwd = 2)
abline(h=New_Mean_Total_Steps, col = "blue", lty = 2, lwd = 2)
```

Activity Patterns for Weekday vs. Weekend.
Create a two parameter factor (DayFactor) for Filled_In

```{r echo = TRUE}
Filled_In$date<-as.Date.character(Filled_In$date)
Filled_In <- mutate(Filled_In, Day = weekdays(date))
DayFactor <- Filled_In$Day
for (i in 1:17568){
  if (DayFactor[i] == "Saturday" | DayFactor[i] == "Sunday"){
    DayFactor[i] <- "Weekend"}
  else{
    DayFactor[i] <- "Weekday"
  }
}
Filled_In <- cbind(Filled_In, DayFactor)
Filled_In <- select(Filled_In, steps, date, Day, DayFactor, interval)
Day_Factor <- Filled_In %>% group_by(DayFactor, interval) %>% 
  summarise(Mean_Steps = mean(steps))
Day_Factor$Mean_Steps <- as.numeric(formatC(Day_Factor$Mean_Steps, digits = 3, format = "f"))
Day_Factor
xyplot(Day_Factor$Mean_Steps ~ Day_Factor$interval | Day_Factor$DayFactor, 
       type = "l", lwd = 1, col = "black", 
       ylab = "Mean Steps", 
       xlab = "Interval",
       layout = c(1,2))
```

